---
title: "Group Assignment"
author: "Jinrui Du S3506401, Thi Kim Oanh Nguyen S3451887"
date: "2024-04-25"
output: pdf_document
---

```{r}
library(survival)
library(tidyverse)
```

```{r}
data = read.table("Oscars.txt", header = T, sep = "\t")

# Verify that each row corresponding to a person.
all(table(data$Identity) == 1)

# Create variable Death from Alive
data$Death = ifelse(data$Alive == 1, 0, 1)

# Create the time to event
data$Age = data$Final - data$Birth

#
data$Oscar_win = ifelse(data$Wins > 0, 1, 0)

#
data$Oscar_nom = ifelse(data$Noms > 0, 1, 0)
```


# Exercise 1, frequncy tables
Make a frequency table of the number of deaths, the number of nominations and wins, and of gender, USA-born and number of films.


```{r}
table(data$Death)
table(data$Noms)
table(data$Wins)
table(data$Male)
table(data$Born_USA)
table(data$Films)
```

# Exercise 2

## 2.1 Overall survival
```{r}
sf_overall = survfit(Surv(Age, Death) ~ 1, data = data)

plot(sf_overall)
```
```{r}
tail(data[order(data$Age), ])

tab = summary(sf_overall)
t = tab$time
t = c(0, t[-length(t)])
cumsum(tab$surv * (tab$time - t))

mus = numeric(length(sf_overall$time))
mu = sf_overall$time[1]
mus[1] = mu
for(i in 1:(length(sf_overall$time)-1)) {
  mu <- mu + sf_overall$surv[i] * (sf_overall$time[i+1] - sf_overall$time[i])
  mus[i + 1] = mu
}
summary(sf_overall)$table["rmean"]

print(sf_overall, print.rmean=TRUE, rmean=120)
```

### Log-rank test
```{r}
survdiff(Surv(Age, Death) ~ Male, data = data)
```
### Cox proportional hazards model
```{r}
coxph(Surv(Age, Death) ~ Male, data = data)
```

# Exercise 3

## 2.1 Oscar win

### Kaplan-Meier curve

```{r}
sf_win = survfit(Surv(Age, Death) ~ Oscar_win, data = data)

plot(sf_win)
```

### Log-rank test

```{r}
survdiff(Surv(Age, Death) ~ Oscar_win, data = data)
```

### Cox proportional hazards model

```{r}
coxph(Surv(Age, Death) ~ Oscar_win, data = data)
```


## 2.2 Oscar nomination

### Kaplan-Meier curve

```{r}
sf_nom = survfit(Surv(Age, Death) ~ Oscar_nom, data = data)

plot(sf_nom)
```
### Log-rank test

```{r}
survdiff(Surv(Age, Death) ~ Oscar_nom, data = data)
```

### Cox proportional hazards model

```{r}
coxph(Surv(Age, Death) ~ Oscar_nom, data = data)
```


# Exercise 4


# Exercise 5
```{r}
data$FirstWin_Age = data$FirstWin_Year - data$Birth
data$Time1 = 0
data$Time2 = data$Age

data_temp = data[!is.na(data$FirstWin_Year), ]
data_temp$Time2 = data_temp$FirstWin_Age
data_temp$Oscar_win = 0
data_temp$Death = 0

data[!is.na(data$FirstWin_Year), "Time1"] = data[!is.na(data$FirstWin_Year), "FirstWin_Age"]

data = rbind(data, data_temp)
data = data[order(data$Identity, data$Time1), ]
```


```{r}
data[, c("Identity", "Time1", "Time2", "FirstWin_Age", "Oscar_win", "Death")]
```

## 5.1 
### Kaplan-Meier curve

```{r}
sf_win = survfit(Surv(Age, Death) ~ Oscar_win, data = data)

plot(sf_win)
```

### Log-rank test

```{r}
survdiff(Surv(Age, Death) ~ Oscar_win, data = data)
```

### Cox proportional hazards model

```{r}
coxph(Surv(Time1, Time2, Death) ~ Oscar_win, data = data)
```

## 5.2
```{r}
coxph(Surv(Time1, Time2, Death) ~ Oscar_win + Born_USA + Male, data = data)
```


