---
title: "Group Assignment 6: Oscar"
author: 
- Jinrui Du S3506401
- Thi Kim Oanh Nguyen S3451887
output: pdf_document
df_print: paged
---

# Introduction
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Import packages 
```{r, message=FALSE, warning=FALSE}
library(survival)
library(tidyverse)
library(survminer)
library(gtsummary)
```

# Data Exploration

In this report, we will utilize data related to the Oscar awards. This dataset pertains to the survival of film actors and actresses within the context of the Oscar awards.


```{r}
data = read.table("Oscars.txt", header = T, sep = "\t")

# Verify that each row corresponding to a person.
all(table(data$Identity) == 1)

# Create variable Death from Alive.
data$Death = ifelse(data$Alive == 1, 0, 1)

# Create the time to event.
data$Age = data$Final - data$Birth

# Create binary variable Oscar_win 
data$Oscar_win = ifelse(data$Wins > 0, 1, 0)

# Create binary variable Oscar_nom
data$Oscar_nom = ifelse(data$Noms > 0, 1, 0)

# Summary of data 
summary(data)
```


# Exercise 1, frequncy tables
Make a frequency table of the number of deaths, the number of nominations and wins, and of gender, USA-born and number of films.


```{r}
table(data$Death)
table(data$Noms)
table(data$Wins)
table(data$Male)
table(data$Born_USA)
table(data$Films)
```

# Exercise 2
## 2.1 Overall survival
```{r, message=FALSE, warning=FALSE}
sf_overall = survfit(formula = Surv(Age, Death) ~ 1, data = data)

ggsurvplot(sf_overall,
           data=data,
           xlim = c(0,110),
           pval = FALSE, 
           conf.int = FALSE,
           surv.median.line = "hv", # Specify median survival
           ggtheme = theme_bw(), 
           palette =c("#E7B800", "#2E9FDF"),
           xlab="Age in years") + 
  labs(color = "Groups")
```

### Calculate mean survival time
```{r}
tail(data[order(data$Age), ])
```
The last record in the data set is censored. Therefore, to calculate the mean survival time, we need to choose an upper limit. If we choose the upper limit to be the maximum age, i.e. 104, in the data set, then the mean survival time is $\int_0^{104} \hat S(t) dt = 76.6$. 

```{r}
mu = sf_overall$time[1]
for(i in 1:(length(sf_overall$time)-1)) {
  mu <- mu + sf_overall$surv[i] * (sf_overall$time[i+1] - sf_overall$time[i])
}

summary(sf_overall)$table["rmean"]
```

If we choose the upper limit to be the largest possible age a person can survive, i.e. 120, then the mean survival time is $\int_0^{120} \hat S(t) dt = 76.8$. 

```{r}
print(sf_overall, print.rmean=TRUE, rmean=120)
```



## 2.2 Survival - males and females
### Kaplan-Meier curve
```{r}
sf_gender = survfit(formula = Surv(Age, Death) ~ Male, data = data)
ggsurvplot(sf_gender,
           data = data,
           xlim = c(0,110),
           pval = FALSE, 
           conf.int = FALSE,
           surv.median.line = "hv", # Specify median survival
           ggtheme = theme_bw(), 
           palette = c("#E7B800", "#2E9FDF"),
           xlab="Age in years") + 
  labs(color = "Groups")
```


### Log-rank test
```{r}
survdiff(Surv(Age, Death) ~ Male, data = data)
```
The log-rank test result indicates a significant difference in survival rates between males and females. The p-value of the test is very small (5e-04), indicating sufficient evidence to reject the null hypothesis of no difference between the two groups.

In this case, the chi-square value (chisq) is 12.2 with 1 degree of freedom. This suggests that the difference between the groups is statistically significant.

### Cox proportional hazards model and hazard ratio 

```{r ,message=FALSE, warning=FALSE}
model = coxph(Surv(Age, Death) ~ Male, data = data)
tbl_regression(model, exp = TRUE)  # from gtsummary package
```

# Exercise 3

## 3.1 Oscar win
```{r, message=FALSE, warning=FALSE}
sf_win <- survfit(formula=Surv(Age, Death) ~ Oscar_win, data=data)
ggsurvplot(sf_win, 
           data = data,
           xlim = c(0,110),
           pval = FALSE, 
           conf.int = FALSE,
           surv.median.line = "hv", # Specify median survival
           ggtheme = theme_bw(), 
           palette = c("#00008B",'#FF0000'),
           xlab="Age in years") + 
  labs(color = "Groups")
```

### Log-rank test
```{r}
survdiff(Surv(Age, Death) ~ Oscar_win, data = data)
```

### Cox proportional hazards model and hazard ratio
```{r}
model2 = coxph(Surv(Age, Death) ~ Oscar_win, data = data)
tbl_regression(model2, exp = TRUE) # exp = TRUE, create a regression table summarizing the Cox regression model
```

At any given time points the actor/ actress with Oscar awards is 0.75 time as likely to die as without Oscar awards

## 3.2 Oscar nomination
```{r, warning=FALSE, message=FALSE}
sf_nom = survfit(Surv(Age, Death) ~ Oscar_nom, data = data)
ggsurvplot(sf_nom,
           data = data,
           xlim = c(0,110),
           pval = FALSE, conf.int = FALSE,
           surv.median.line = "hv", # Specify median survival
           ggtheme = theme_bw(), 
           palette =c( "#2E9FDF", "#E7B800"),
           xlab="Age in Years") + 
  labs(color = "Groups")
```
### Log-rank test
```{r}
survdiff(Surv(Age, Death) ~ Oscar_nom, data = data)
```

### Cox proportional hazards model
```{r}
model3 = coxph(Surv(Age, Death) ~ Oscar_nom, data = data)
tbl_regression(model3, exp = TRUE) 
```
At any given time points the actor/ actress with nomination is 0.87 time as likely to die as without nomination 


# Exercise 4: 
What is your conclusion? Do a similar analysis
comparing Oscar nominees with actors/actresses without Oscar nomination.

# Exercise 5
```{r}
data$FirstWin_Age = data$FirstWin_Year - data$Birth
data$Time1 = 0
data$Time2 = data$Age

data_temp = data[!is.na(data$FirstWin_Year), ]
data_temp$Time2 = data_temp$FirstWin_Age
data_temp$Oscar_win = 0
data_temp$Death = 0

data[!is.na(data$FirstWin_Year), "Time1"] = data[!is.na(data$FirstWin_Year), "FirstWin_Age"]

data = rbind(data, data_temp)
data = data[order(data$Identity, data$Time1), ]
```


```{r}
data[, c("Identity", "Time1", "Time2", "FirstWin_Age", "Oscar_win", "Death")]
```

## 5.1 
### Kaplan-Meier curve
```{r}
sf_win = survfit(Surv(Age, Death) ~ Oscar_win, data = data)

plot(sf_win)
```

### Log-rank test
```{r}
survdiff(Surv(Age, Death) ~ Oscar_win, data = data)
```

### Cox proportional hazards model
```{r}
coxph(Surv(Time1, Time2, Death) ~ Oscar_win, data = data)
```

## 5.2
```{r}
coxph(Surv(Time1, Time2, Death) ~ Oscar_win + Born_USA + Male, data = data)
```






