---
title: "Assignment 6: Oscar data"
author: 
- Jinrui Du S3506401, Thi Kim Oanh Nguyen S3451887
output: pdf_document
df_print: paged
---

# Introduction

In this report, we aim to investigate the effect of winning an Oscar on the survival of actors and actresses. Building upon the work of Redelmeier and Singh (2001), who concluded that Oscar wins prolonged survival, we aim to incorporate Oscar wins as a time-dependent variable. Our analysis involves data stratification, Kaplan-Meier estimation, log-rank tests, and Cox proportional hazards modeling. Specifically, we explore survival patterns among actors, considering factors such as gender and age.

# Import packages

```{r, message=FALSE, warning=FALSE}
rm(list = ls())
library(survival)
library(tidyverse)
library(survminer)
library(gtsummary)
```

# 1. Data exploration

The Oscar data originated from the aforementioned analysis. 

```{r}
data = read.table("Oscars.txt", header = T, sep = "\t")

# Verify that each row corresponding to a person.
all(table(data$Identity) == 1)

# Create variable Death from Alive.
data$Death = ifelse(data$Alive == 1, 0, 1)

# Create the time to event.
data$Age = data$Final - data$Birth

# Create binary variable Oscar_win 
data$Oscar_win = ifelse(data$Wins > 0, 1, 0)

# Create binary variable Oscar_nom
data$Oscar_nom = ifelse(data$Noms > 0, 1, 0)

# Summary of data 
summary(data)
```

The data has information of 1670 actors spanned across 11 variables. It contains their year of birth, year of death or last known to be alive (i.e. 2001), from the two we get their age. We also get an indicator variable showing if the death was censored. The year of first Oscar win is also in the data set. Other variables are the number of films, born in USA, gender, number of Oscar nomination, number of Oscar wins, and the year of first Oscar nomination. Our goal is to understand the survival patterns among these actors and to determine if factors like gender influence survival. We will employ statistical techniques including Kaplan-Meier survival analysis and Cox proportional hazards modeling. Below, we present the initial analysis:


## Frequency table of the number of deaths

```{r}
death_table = table(data$Death)
death_df = as.data.frame(death_table)
names(death_df) = c("Death", "Frequency")
ggplot(death_df, aes(x = Death, y = Frequency, fill = as.factor(Death))) +
  geom_bar(stat = "identity") +
    labs(title = "Frequency of Deaths") +
  geom_text(aes(label = Frequency), vjust = -0.5, color = "red", size = 4) + # Add count values
  scale_fill_manual(values = c("lightblue", "grey"), labels = c("Alive", "Dead")) +
  scale_x_discrete(labels = c("0" = "Alive", "1" = "Dead")) +
  theme(legend.position = "none")
```

The number of dead and alive actors up to 2001 is nearly balanced, with a difference of only 92.


## Frequency table of the number of nominations

```{r}
nom_table = table(data$Noms)
nom_df = as.data.frame(nom_table)
names(nom_df) = c("Nomination", "Frequency")
ggplot(nom_df, aes(x = Nomination, y = Frequency, fill = as.factor(Nomination))) +
  geom_bar(stat = "identity") +
    labs(title = "Frequency of Nomination", x = "Number of Nominations") +
  geom_text(aes(label = Frequency), vjust = -0.5, color = "red", size = 4) +
  theme(legend.position = "none")
```

The frequency decreases as the number of nominations increases. Most actors (902) have not received any Oscar nominations and 496 actors have received one nomination. Only two actors have received eleven nominations, and another two have received twelve nominations, which is the highest number of nominations.


## Frequency table of the number of wins

```{r}
win_table = table(data$Wins)
win_df = as.data.frame(win_table)
names(win_df) = c("Wining_Oscar", "Frequency")
ggplot(win_df, aes(x = Wining_Oscar, y = Frequency, fill = as.factor(Wining_Oscar))) +
  geom_bar(stat = "identity") +
    labs(title = "Frequency of  Wining Oscar", x = "Number of Wins") +
  geom_text(aes(label = Frequency), vjust = -0.5, color = "red", size = 4) +
  theme(legend.position = "none")
```

There are 1670 observations, but 1431 of them have zero Oscar wins, highlighting a significant imbalance in the dataset when comparing different groups based on the number of wins. Notably, there is one actor who has won the Oscar four times.


## Frequency table of gender

```{r}
gender_table = table(data$Male)
gender_df = as.data.frame(gender_table)
names(gender_df) = c("Gender", "Frequency")
ggplot(gender_df, aes(x = Gender, y = Frequency, fill = as.factor(Gender))) +
  geom_bar(stat = "identity") +
    labs(title = "Frequency of Gender") +
  geom_text(aes(label = Frequency), vjust = -0.5, color = "red", size = 4) + # Add count values
  scale_fill_manual(values = c("pink","lightblue"), labels = c("Female", "Male")) +
  scale_x_discrete(labels = c("0" = "Female", "1" = "Male")) +
  theme(legend.position = "none")
```
The difference in the numbers of male and female actors is not significant. There are more male actors (893) than female actors (777). 


## Frequency table of USA born

```{r}
born_table = table(data$Born_USA)
born_df = as.data.frame(born_table)
names(born_df) = c("Born_USA", "Frequency")
ggplot(born_df, aes(x = Born_USA, y = Frequency, fill = as.factor(Born_USA))) +
  geom_bar(stat = "identity") +
    labs(title = "Frequency of Born in USA", x = "Born in USA") +
  geom_text(aes(label = Frequency), vjust = -0.5, color = "red", size = 4) + # Add count values
  scale_fill_manual(values = c("darkgreen", "orange"), labels = c("No", "Yes")) +
  scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) + 
  theme(legend.position = "none")
```

There are 1,193 actors born in the USA and 477 actors born elsewhere. The number of US-born actors is approximately 2.5 times that of non-US-born actors in the data set.
The dataset is imbalanced in terms of birthplace.


## Frequency table of number of films
```{r}
hist(data$Films,
     breaks = seq(0, 300, by = 10), 
     main = "Distribution of Film Counts", 
     xlab = "Film Counts", 
     ylab = "Frequency")
```

Indeed, it is quite surprising to see actors with such a high number of films in his/her lifetime. The histogram indicates that while most actors have appeared in fewer than 50 films, there are outliers who have been in over 250 films. This demonstrates the significant variation in the careers of actors in the dataset.


# 2. Kaplan-Meier curve for overall survival, mean survival time, and comparing survival of actors and actresses.

## 2.1 Overall survival

We plot the Kaplan-Meier curve for all the actors in the data set.

```{r, message=FALSE, warning=FALSE}
sf_overall = survfit(formula = Surv(Age, Death) ~ 1, data = data)

ggsurvplot(sf_overall,
           data=data,
           xlim = c(0,110),
           pval = FALSE, 
           conf.int = FALSE,
           surv.median.line = "hv", # Specify median survival
           ggtheme = theme_bw(), 
           palette =c("#E7B800", "#2E9FDF"),
           xlab="Age in years") + 
  labs(color = "Groups")

summary(sf_overall)$table["median"]
```

The Kaplan-Meier survival curve for all actors in the dataset shows a relatively stable survival rate from early adulthood through middle age, with a noticeable decline beginning around age 50. This pattern suggests that actors tend to have consistent survival rates in their earlier years, followed by a gradual decrease in survival rates as they age. The median survival age, indicating the age at which 50% of actors have passed away, is estimated to be 79 years. This analysis provides insight into the overall longevity trends among actors in the dataset.


### Calculate Mean Survival Time

To determine the mean survival time of actors in the dataset, we first assess if the last entry is censored:

```{r}
tail(data[order(data$Age), ])
```

The last record in the dataset is censored. Consequently, we calculate the mean survival time by choosing an upper limit. Setting the upper limit at the maximum age observed in the dataset, 104 years, yields a mean survival time of approximately 76.6 years:

 $\int_0^{104} \hat S(t) dt = 76.6$.

```{r}
mu = sf_overall$time[1]
for(i in 1:(length(sf_overall$time)-1)) {
  mu <- mu + sf_overall$surv[i] * (sf_overall$time[i+1] - sf_overall$time[i])
}

summary(sf_overall)$table["rmean"]
```

Alternatively, if we extend the upper limit to 120 years, assuming it as the maximum potential age of human, the mean survival time increases slightly to 76.8 years:

 $\int_0^{120} \hat S(t) dt = 76.8$.

```{r}
print(sf_overall, print.rmean=TRUE, rmean=120)
```

These calculations provide insight into the average survival duration of actors in the dataset, accounting for both the observed data and potential extended lifespans.


## 2.2 Survival - males and females

Next, we’ll compare the survival curves with Kaplan-Meier curves between male and female actors. This analysis will help us understand whether gender plays a significant role in survival.

### Kaplan-Meier curve

```{r, warning=FALSE, message=FALSE}
sf_gender = survfit(formula = Surv(Age, Death) ~ Male, data = data)
ggsurvplot(sf_gender,
           data = data,
           xlim = c(0,110),
           pval = FALSE, 
           conf.int = FALSE,
           surv.median.line = "hv", # Specify median survival
           ggtheme = theme_bw(), 
           palette = c("#E7B800", "#2E9FDF"),
           xlab="Age in years") + 
  labs(color = "Groups")
```
```{r}
summary(sf_gender)$table
```

The Kaplan-Meier curves highlight distinct survival trends between males (blue curve) and females (yellow curve), with females consistently showing a higher survival rate across all ages starting from around 62.5 to 93 years. For females, with 777 records, the median survival age is 81 years, with a 95% confidence interval (CI) ranging from 80 to 83 years. The estimated mean survival time is around 78.06 years. In comparison, males, with 893 records, have a median survival age of 77 years, with a 95% CI ranging from 76 to 78 years, and an estimated mean survival time of approximately 75.5 years. The observed higher survival rates among females compared to males align with broader demographic trends, highlighting gender as a significant factor influencing survival outcomes in our dataset.


### Log-rank test

We use the log-rank test to assesses whether there is a statistically significant difference in survival between male and female actors.

```{r}
survdiff(Surv(Age, Death) ~ Male, data = data)
```
The log-rank test result shows that the test statistic (Chi-squared) is 12.2 with 1 degree of freedom, and the p-value is very small (5e-04). We reject the null hypothesis, indicating that there is indeed a significant difference in survival between the two gender groups. Therefore, gender plays a crucial role in influencing actors’ survival outcomes.


### Cox proportional hazards model and hazard ratio

```{r ,message=FALSE, warning=FALSE}
model = coxph(Surv(Age, Death) ~ Male, data = data)
tbl_regression(model, exp = TRUE)
```

The Cox model estimates the hazard ratio (HR) associated with gender (Male). The hazard ratio represents the relative risk of death for one group compared to another. For male actors, the hazard ratio is 1.29 (95% CI: 1.12 to 1.48), with a p-value lower 0.001. Thus, male actors have a 29% higher risk of death compared to female actors at a given age. 

In summary, gender significantly influences survival among actors, with males facing a higher mortality risk.


# 3. Oscar win and nomination as time-fixed variables

## 3.1 Oscar win

```{r, message=FALSE, warning=FALSE}
sf_win <- survfit(formula=Surv(Age, Death) ~ Oscar_win, data=data)
ggsurvplot(sf_win, 
           data = data,
           xlim = c(0,110),
           pval = FALSE, 
           conf.int = FALSE,
           surv.median.line = "hv", # Specify median survival
           ggtheme = theme_bw(), 
           palette = c("#00008B",'#FF0000'),
           xlab="Age in years") + 
  labs(color = "Groups")
```

### Log-rank test

```{r}
survdiff(Surv(Age, Death) ~ Oscar_win, data = data)
```

### Cox proportional hazards model and hazard ratio

```{r}
model2 = coxph(Surv(Age, Death) ~ Oscar_win, data = data)
tbl_regression(model2, exp = TRUE) # exp = TRUE, create a regression table summarizing the Cox regression model
```

At any given time points the actor/ actress with Oscar awards is 0.75 time as likely to die as without Oscar awards

## 3.2 Oscar nomination

```{r, warning=FALSE, message=FALSE}
sf_nom = survfit(Surv(Age, Death) ~ Oscar_nom, data = data)
ggsurvplot(sf_nom,
           data = data,
           xlim = c(0,110),
           pval = FALSE, conf.int = FALSE,
           surv.median.line = "hv", # Specify median survival
           ggtheme = theme_bw(), 
           palette =c( "#2E9FDF", "#E7B800"),
           xlab="Age in Years") + 
  labs(color = "Groups")
```

### Log-rank test

```{r}
survdiff(Surv(Age, Death) ~ Oscar_nom, data = data)
```

### Cox proportional hazards model

```{r}
model3 = coxph(Surv(Age, Death) ~ Oscar_nom, data = data)
tbl_regression(model3, exp = TRUE) 
```

At any given time points the actor/ actress with nomination is 0.87 time as likely to die as without nomination

# 4. The reason that above analysis is wrong.

# 5. Oscar win and nomination as time-dependent variables


```{r}
data$FirstWin_Age = data$FirstWin_Year - data$Birth
data$Time1 = 0
data$Time2 = data$Age

data_temp = data[!is.na(data$FirstWin_Year), ]
data_temp$Time2 = data_temp$FirstWin_Age
data_temp$Oscar_win = 0
data_temp$Death = 0

data[!is.na(data$FirstWin_Year), "Time1"] = data[!is.na(data$FirstWin_Year), "FirstWin_Age"]

data = rbind(data, data_temp)
data = data[order(data$Identity, data$Time1), ]
```

```{r}
data[11:15, c("Identity", "Time1", "Time2", "FirstWin_Age", "Oscar_win", "Death")]
```

## 5.1 Single variate Cox regression

### Kaplan-Meier curve

```{r, warning=FALSE, message=FALSE}
sf_win = survfit(Surv(Age, Death) ~ Oscar_win, data = data)
ggsurvplot(sf_win, 
           data = data,
           xlim = c(0,110),
           pval = FALSE, 
           conf.int = FALSE,
           surv.median.line = "hv", # Specify median survival
           ggtheme = theme_bw(), 
           palette = c("#00008B",'#FF0000'),
           xlab="Age in years") + 
  labs(color = "Groups")
```

### Log-rank test

```{r}
survdiff(Surv(Age, Death) ~ Oscar_win, data = data)
```

### Cox proportional hazards model

```{r}
model4 = coxph(Surv(Time1, Time2, Death) ~ Oscar_win, data = data)
tbl_regression(model4, exp = TRUE) 
```

## 5.2 Multivariate Cox regression

```{r}
model5 = coxph(Surv(Time1, Time2, Death) ~ Oscar_win + Born_USA + Male, data = data)
tbl_regression(model5, exp = TRUE) 
```

