---
title: "Group Assignment 3: Oscar"
author: 
- Jinrui Du S3506401
- Thi Kim Oanh Nguyen S3451887
output: pdf_document
df_print: paged
---

# Introduction



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Import packages 
```{r, message=FALSE, warning=FALSE}
library(survival)
library(tidyverse)
library(survminer)
library(gtsummary)
```

# Data Exploration

In this report, we will utilize data related to the Oscar awards. This dataset pertains to the survival of film actors and actresses within the context of the Oscar awards.

```{r}
data = read.table("Oscars.txt", header = T, sep = "\t")
# Verify that each row corresponding to a person.
all(table(data$Identity) == 1)
# Create variable Death from Alive
data$Death = ifelse(data$Alive == 1, 0, 1)
# Create the time to event
data$Year = data$Final - data$Birth
summary((data))
```


# Exercise 1, frequncy tables
Make a frequency table of the number of deaths, the number of nominations and wins, and of gender, USA-born and number of films.


```{r}
table(data$Death)
table(data$Noms)
table(data$Wins)
table(data$Male)
table(data$Born_USA)
table(data$Films)
```

# Exercise 2
## 2.1 Overall survival
```{r, message=FALSE, warning=FALSE}
sf_overall <- survfit(formula=Surv(Year,Death) ~ 1, data=data)
ggsurvplot(sf_overall,data=data,
# from survminer package
           xlim = c(0,110),
           pval = FALSE, conf.int = FALSE,
           surv.median.line = "hv", # Specify median survival
           ggtheme = theme_bw(), 
           palette =c("#E7B800", "#2E9FDF") ,xlab="Years") + labs(color = "Groups")

```

```{r}
# Calculate the mean survival time
print(sf_overall, print.rmean=TRUE, rmean=120)
```

```{r}
summary(sf_overall)$table[5:6]
```



## 2.2 Survival - males and females
### Kaplan-Meier curve
```{r}

sf_gender <- survfit(formula=Surv(Year,Death) ~ Male, data=data)
ggsurvplot(sf_gender,data=data,
           xlim = c(0,110),
           pval = FALSE, conf.int = FALSE,
           surv.median.line = "hv", # Specify median survival
           ggtheme = theme_bw(), 
           palette =c("#E7B800", "#2E9FDF"),,xlab="Years") + labs(color = "Groups")

```





### Log-rank test
```{r}
survdiff(Surv(Year, Death) ~ Male, data = data)
```
The log-rank test result indicates a significant difference in survival rates between males and females. The p-value of the test is very small (5e-04), indicating sufficient evidence to reject the null hypothesis of no difference between the two groups.

In this case, the chi-square value (chisq) is 12.2 with 1 degree of freedom. This suggests that the difference between the groups is statistically significant.

### Cox proportional hazards model
### Hazard ratio 

```{r ,message=FALSE, warning=FALSE}
model <- coxph(Surv(Year, Death) ~ Male, data = data)
tbl_regression(model,exp = TRUE)  # from gtsummary package
```

# Exercise 3

```{r, message=FALSE, warning=FALSE}
#  Oscar win
data$Oscar_win = ifelse(data$Wins > 0, 1, 0)
sf_gender <- survfit(formula=Surv(Year,Death) ~ Male, data=data)
ggsurvplot(sf_win,data=data,
           xlim = c(0,110),
           pval = FALSE, conf.int = FALSE,
           surv.median.line = "hv", # Specify median survival
           ggtheme = theme_bw(), 
           palette =c("#00008B",'#FF0000'),xlab="Years") + labs(color = "Groups")
```

### Log-rank test
### Cox proportional hazards model
```{r}
survdiff(Surv(Year, Death) ~ Oscar_win, data = data)
model2 = coxph(Surv(Year, Death) ~ Oscar_win, data = data)
tbl_regression(model2, exp = TRUE) 
# exp = TRUE , create a regression table summarizing the Cox regression model
```

At any given time points the actor/ actress with Oscar awards is 0.75 time as likely to die as without Oscar awards
```{r, warning=FALSE, message=FALSE}
## Oscar nomination
data$Oscar_nom = ifelse(data$Noms > 0, 1, 0)
sf_nom = survfit(Surv(Year, Death) ~ Oscar_nom, data = data)
ggsurvplot(sf_nom,data=data,
           xlim = c(0,110),
           pval = FALSE, conf.int = FALSE,
           surv.median.line = "hv", # Specify median survival
           ggtheme = theme_bw(), 
           palette =c( "#2E9FDF", "#E7B800"),xlab="Years") + labs(color = "Groups")
```
### Log-rank test
### Cox proportional hazards model

```{r}
survdiff(Surv(Year, Death) ~ Oscar_nom, data = data)
model3 = coxph(Surv(Year, Death) ~ Oscar_nom, data = data)
tbl_regression(model3, exp = TRUE) 
```
At any given time points the actor/ actress with nomination is 0.87 time as likely to die as without nomination 


Question 4: 
What is your conclusion? Do a similar analysis
comparing Oscar nominees with actors/actresses without Oscar nomination.










